From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-01 Pi+/paper/reversible1.tex
denotation 1

%%%%%%%%%
\subsection{Denotational Semantics (Outline)}

\noindent The accompanying code includes an operational semantics for $\Pi$. A denotational semantics can be
specified by mapping each type to a finite set and each combinator to a bijection between finite sets, that is, a
permutation. In this section, we outline a simple semantics expressed in the metalanguage of sets and functions. This
semantics will be formalized and extended to the metalanguage of homotopy type theory in Sec.~\ref{sec:reversibletwo}.

As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
$\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
\[\begin{array}{rcl}
    \denot{\onet + (\onet + \onet)} &=& \{ 0,1,2 \} \\
    \denot{(\onet + \onet) + \onet} &=& \{ 0,1,2 \} \\
    \\
    \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (0~1~2) \\
    \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (2~0~1)
\end{array}\]
where we have used the one-line notation for permutations with $(a~b~c)$ representing the
permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
enumeration of the elements of each type:
\[\begin{array}{rcl}
    \mathit{enum}(\zerot) &=& [~] \\
    \mathit{enum}(\onet) &=& [ ~\Acon{tt}~ ] \\
    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B) \\
    \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
\end{array}\]
\noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
operation that appends all the subsequences in a sequence.

Using the definition, we have:
\[\begin{array}{rcl}
    \mathit{enum}(\onet + (\onet + \onet)) &=& [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
    \mathit{enum}((\onet + \onet) + \onet) &=& [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
\end{array}\]
Thus, as shown in the diagrams below, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of
$\inlv{\Acon{tt}}$ which is 0 and produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value
$\inlv{(\inlv{\Acon{tt}})}$. Similarly, $\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of
$\inlv{\Acon{tt}}$ which is 0 and produces index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value
$\inrv{\Acon{tt}}$.

\begin{center}
  \input{assoc-perm.tikz}
  \qquad
  \input{swap-perm.tikz}
\end{center}

% Such functions are finitely supported, that is, their outputs can be tabulated using the canonical ordering on
% $\Fin[n]$. For bijections, this gives a listed permutation. By observing the action of the combinators on the values of
% the finite set, we can define a denotational semantics which constructs the bijection.
% \note{\(\Fin[n] \to A \eqv \type{Vec_{n}}(A)\)}

%%%%%%%%%
\subsection{Examples}
\label{sec:langRev-examples}
\label{examples}

@@

\noindent The input value of type $\mathbb{2} \times A$ is processed by the distribute operator \ensuremath{\dist},
which converts it into a value of type $(\onet \times A) + (\onet \times A)$. In the \ensuremath{\mathit{left}} branch,
which corresponds to the case when the boolean is \ensuremath{\mathit{false}}, the combinator~\ensuremath{c_1} is
applied to the value of type~\ensuremath{A}. The right branch which corresponds to the boolean being
\ensuremath{\mathit{true}} passes the value of type $A$ through the combinator \ensuremath{c_2}.  The inverse of
\ensuremath{\dist}, namely \ensuremath{\factor} is applied to get the final result. The full expressiveness of $\Pi$ is
demonstrated by the following implementation of $\Afun{cnot}$ and $\Afun{toffoli}$ which are universal for reversible
computation~\cite{Toffoli:1980}. The $\Afun{cnot}$ gates takes two input bits: a control bit and a target bit, and
negates the target bit if the control bit is set. It is essentially a one-branch version of $\Afun{cif}$ above. The
Toffoli gate generalizes the idea to a gate that takes 3 input bits: 2 control bits and one target bit: boolean inputs:
it negates the target bit if both control bits are set.

>>>>

%%%%%%%%%
\subsection{Informal Semantics and Examples}
\label{sec:langRev-examples}
\label{examples}

@@

\noindent The input value of type $\mathbb{2} \times A$ is processed by the distribute operator \ensuremath{\dist},
which converts it into a value of type $(\onet \times A) + (\onet \times A)$. In the \ensuremath{\mathit{left}} branch,
which corresponds to the case when the boolean is \ensuremath{\mathit{false}}, the combinator~\ensuremath{c_1} is
applied to the value of type~\ensuremath{A}. The right branch which corresponds to the boolean being
\ensuremath{\mathit{true}} passes the value of type $A$ through the combinator \ensuremath{c_2}.  The inverse of
\ensuremath{\dist}, namely \ensuremath{\factor} is applied to get the final result. The full expressiveness of $\Pi$ is
demonstrated by an implementation of $\Afun{cnot}$ and $\Afun{toffoli}$ which are universal for reversible
computation~\cite{Toffoli:1980}. The $\Afun{cnot}$ gates takes two input bits: a control bit and a target bit, and
negates the target bit if the control bit is set. It is essentially a one-branch version of $\Afun{cif}$ above. The
Toffoli gate generalizes the idea to a gate that takes 3 input bits: 2 control bits and one target bit: boolean inputs:
it negates the target bit if both control bits are set.

%%%%%%%%%
\subsection{Denotational Semantics}

\noindent The accompanying code includes an operational semantics for $\Pi$. A denotational semantics can be
specified by mapping each type to a finite set and each combinator to a bijection between finite sets, that is, a
permutation. In this section, we outline a simple semantics expressed in the metalanguage of sets and functions. This
semantics will be extended to the metalanguage of homotopy type theory in the remainder of the paper.

The denotation of types is straightforward:
\[\begin{array}{rcl}
\denot{\zerot} &=& \emptyset \\
\denot{\onet} &=& \{ \Acon{tt} \} \\
\denot{A + B} &=& \denot{A} \uplus \denot{B} \\
\denot{A \times B} &=& \denot{A} \times \denot{B}
\end{array}\]
where $\emptyset$ is the empty set, $\uplus$ is the disjoint union of sets, and $\times$ is the cartesian product of
sets. The denotation of a combinator $c : A \isot B$ is a (bijective) function mapping $\denot{A}$ to $\denot{B}$:
\[\begin{array}{rcl}
\denot{\identlp} (\inlv{v}) &=& v \\
\denot{\identrp} v &=& \inlv{v} \\
\denot{\swapp} (\inlv{v}) &=& \inrv{v} \\
\denot{\swapp} (\inrv{v}) &=& \inlv{v} \\
\denot{\assoclp} (\inlv{v})  &=& \inlv{(\inlv{v})} \\
\denot{\assoclp} (\inrv{(\inlv{v})}) &=& \inlv{(\inrv{v})} \\
\denot{\assoclp} (\inrv{(\inrv{v})}) &=& \inrv{v} \\
\denot{\assocrp} (\inlv{(\inlv{v})}) &=& \inlv{v} \\
\denot{\assocrp} (\inlv{(\inrv{v})}) &=& \inrv{(\inlv{v})} \\
\denot{\assocrp} (\inrv{v})  &=& \inrv{(\inrv{v})} \\
\denot{\identlp} (\Acon{tt} , v)  &=& v \\
\denot{\identrp} v &=& (\Acon{tt} , v) \\
\denot{\swapt}  (v_1 , v_2) &=& (v_2 , v_1) \\
\denot{\assoclt} (v_1 , (v_2 , v_3)) &=& ((v_1 , v_2) , v_3) \\
\denot{\assocrt} ((v_1 , v_2) , v_3) &=& (v_1 , (v_2 , v_3)) \\
\denot{\dist} (\inlv{v_1} , v_3)  &=& \inlv{(v_1 , v_3)} \\
\denot{\dist} (\inrv{v_2 , v_3})  &=& \inrv{(v_2 , v_3)} \\
\denot{\factor}  (\inlv{(v_1 , v_3)}) &=& (\inlv{v_1} , v_3) \\
\denot{\factor}  (\inrv{(v_2 , v_3)}) &=& (\inrv{v_2} , v_3) \\
\denot{\idc} v &=& v \\
\denot{(c_1 \fatsemi c_2)} v &=& (\denot{c_2} \circ \denot{c_1}) v \\
\denot{(c_1 \oplus c_2)} (\inlv{v}) &=& \inlv{(\denot{c_1}~v)} \\
\denot{(c_1 \oplus c_2)} (\inrv{v}) &=& \inrv{(\denot{c_2}~v)} \\
\denot{(c_1 \otimes c_2)} (v_1 , v_2) &=& (\denot{c_1} v_1 , \denot{c_2} v_2)
  \end{array}\]

% \begin{center}
% \begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
%   \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
%   \draw[fill] (-2,-1.5) circle [radius=0.025];
%   \node[below] at (-2.1,-1.5) {$A$};
%   \node[below] at (-2.1,-1.9) {$+$};
%   \draw[fill] (-2,-2.5) circle [radius=0.025];
%   \node[below] at (-2.1,-2.5) {$B$};

%   \draw[fill] (6.5,-1.5) circle [radius=0.025];
%   \node[below] at (6.7,-1.5) {$C$};
%   \node[below] at (6.7,-1.9) {$+$};
%   \draw[fill] (6.5,-2.5) circle [radius=0.025];
%   \node[below] at (6.7,-2.5) {$D$};

%   \draw[<-] (-2,-1.5) to[bend left] (1,0.5) ;
%   \draw[<-] (-2,-2.5) to[bend left] (1,-0.5) ;
%   \draw[->] (3.5,0.5) to[bend left] (6.5,-1.45) ;
%   \draw[->] (3.5,-0.5) to[bend left] (6.5,-2.45) ;

%   \draw[<-] (-2,-1.5) to[bend right] (1,-3.5) ;
%   \draw[<-] (-2,-2.5) to[bend right] (1,-4.5) ;
%   \draw[->] (3.5,-3.5) to[bend right] (6.5,-1.55) ;
%   \draw[->] (3.5,-4.5) to[bend right] (6.5,-2.55) ;

%   \draw     (2,0.5)  -- (2.5,0.5)  ;
%   \draw     (2,-0.5) -- (2.5,-0.5) ;

%   \draw     (2.5,0.5)  -- (3.5,-0.5)  ;
%   \draw     (2.5,-0.5) -- (3.5,0.5) ;

%   \draw     (1,-3.5)  -- (2,-4.5)    ;
%   \draw     (1,-4.5) -- (2,-3.5)   ;

%   \draw     (2,-3.5)  -- (2.5,-3.5)    ;
%   \draw     (2,-4.5) -- (2.5,-4.5)   ;

%   \path (1.5,0.5) node (tc1) [func] {$c_1$};
%   \path (1.5,-0.5) node (tc2) [func] {$c_2$};
%   \path (3,-4.5) node (bc1) [func] {$c_1$};
%   \path (3,-3.5) node (bc2) [func] {$c_2$};
% \end{tikzpicture}
% \end{center}
% The top path is the $\Pi$ program $(c_1~\oplus~c_2)~\odot~\swapp$ which acts on the type $A$ by $c_1$, acts on the type
% $B$ by $c_2$, and acts on the resulting value by a twist that exchanges the two injections into the sum type. The bottom
% path performs the twist first and then acts on the type $A$ by $c_1$ and on the type $B$ by $c_2$ as before. One could
% imagine the paths are physical \emph{elastic} wires in $3$ space, where the programs $c_1$ and $c_2$ as arbitrary
% deformations on these wires, and the twists do not touch but are in fact well-separated. Then, holding the points $A$,
% $B$, $C$, and $D$ fixed, it is possible to imagine sliding $c_1$ and $c_2$ from the top wire rightward past the twist,
% and then using the elasticity of the wires, pull the twist back to line up with that of the bottom --- thus making both
% parts of the diagram identical.  Each of these moves can be undone (reversed), and doing so would take the bottom part
% of the diagram into the top part.  In other words, there exists an equivalence of the program
% $(c_1~\oplus~c_2)~\odot~\swapp$ to the program $\swapp \odot (c_2~\oplus~c_1)$. We can also show that this means that,
% as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and $\swapp \odot (c_2~\oplus~c_1)$ are equal. And, of course, not all
% programs between the same types can be deformed into one another. The simplest example of inequivalent deformations are
% the two automorphisms of $1+1$, namely $\idc$ and $\swapp$.

The denotational semantics can be used to evaluate programs and to check equivalence of programs. For example, it is
straightforward to verify that $\denot{(c_1~\oplus~c_2)~\fatsemi~\swapp} = \denot{\swapp \fatsemi (c_2~\oplus~c_1)}$. A
slightly more involved example are the following two programs:

\rotate{}

\noindent The first program performs the following sequence of transformations:
\[
\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
\Tree [ [ {\small a} {\small b} ] {\small c} ] ~\to~
\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
\]
\noindent
while the second evaluates as follows:
\[
\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
\Tree [ {\small a} [ {\small c} {\small b} ] ] ~\to~
\Tree [ [ {\small a} {\small c} ] {\small b} ] ~\to~
\Tree [ [ {\small c} {\small a} ] {\small b} ] ~\to~
\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
\]

The semantics above can be packaged in the category of finite sets and functions, $\SetFin$, which is the category
freely generated by finite coproduct completion of the terminal category. Objects of $\SetFin$ can be identified with
sets of fixed cardinality, that is, $\Fin[n] \defeq \Set{0,1,\ldots,n-1}$. $\SetFin$ has finite coproducts and products,
which lets us interpret the types of $\PiLang$. Combinators are intepreted as morphisms in $\SetFin$, but we have to
restrict to invertible morphisms, that is, isomorphisms. This gives the \emph{groupoid} of finite sets and bijections,
$\BFin \defeq \mathsf{core}(\SetFin)$. The isomorphisms satisfied by coproducts and products in $\SetFin$ lift to
$\BFin$, but they're no longer categorical coproducts and products. They give two symmetric monoidal tensor products on
$\BFin$, the additive and multiplicative ones, with the multiplicative tensor distributing over the additive tensor.

\begin{theorem}\label{thm:grp}
  Let $\sim$ be the following equivalence relation on combinators $c_1 \sim c_2$ iff $\denot{c_1} = \denot{c_2}$
  identifying combinators with the same denotation.  The notation $[c]_{\sim}$ refers to a representative combinator
  from a $\sim$-equivalence class. Using this relation, we define the
  category $\mathcal{C}$ as follows:
  \begin{itemize}
    \item $\mathit{Obj}(\mathcal{C})$ is the set of $\Pi$-types, and
    \item $\mathit{Hom}(A,B) = [c : A \isot B]_{\sim}$
    \end{itemize}
  The category $\mathcal{C}$ is a groupoid.
\end{theorem}
\begin{proof}
The relation $\sim$ identifies $c$, $\idc \circ c$, and $c \circ \idc$, and is associative. Furthermore, the denotation
of each combinator is a reversible function thus making every morphism into an isomorphism.
\end{proof}

\note{Then say that $\mathcal{C}$ is ``the same'' as $\BFin$ We can say here that the relation $\sim$ also identifies
 $(c_1~\oplus~c_2)~\fatsemi~\swapp$ and $\swapp \fatsemi (c_2~\oplus~c_1)$ and other properties. ?? }

% Such functions are finitely supported, that is, their outputs can be tabulated using the canonical ordering on
% $\Fin[n]$. For bijections, this gives a listed permutation. By observing the action of the combinators on the values of
% the finite set, we can define a denotational semantics which constructs the bijection.
% \note{\(\Fin[n] \to A \eqv <\type{Vec_{n}}(A)\)}

% In the previous section, we examined equivalences between conventional data structures, i.e., structured trees of
% values. We now consider a richer but foundational notion of data: programs themselves. Indeed, universal computation
% models crucially rely on the fact that \emph{programs are (or can be encoded as) data}, e.g., a Turing machine can be
% encoded as a string that another Turing machine (or even the same machine) can manipulate. Similarly, first-class
% functions are the \emph{only} values in the $\lambda$-calculus.  In our setting, we ask whether the programs developed
% in the previous section can themselves be subject to (higher-level) equivalences?

### 2021-07-01 Pi+/paper/ufin.tex
denotation 1

>>>>

%%%
\subsection{To Put Somewhere or Delete}

This approach is sufficient to prove the semantics forms a 1-category but ignores the rich structure at the next
level~\cite{DBLP:conf/esop/CaretteS16}.
As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
$\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
\[\begin{array}{rcl}
    \denot{\onet + (\onet + \onet)} &=& \{ 0,1,2 \} \\
    \denot{(\onet + \onet) + \onet} &=& \{ 0,1,2 \} \\
    \\
    \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (0~1~2) \\
    \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (2~0~1)
\end{array}\]
where we have used the one-line notation for permutations with $(a~b~c)$ representing the
permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
enumeration of the elements of each type:
\[\begin{array}{rcl}
    \mathit{enum}(\zerot) &=& [~] \\
    \mathit{enum}(\onet) &=& [ ~\Acon{tt}~ ] \\
    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B) \\
    \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
\end{array}\]
\noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
operation that appends all the subsequences in a sequence.

Using the definition, we have:
\[\begin{array}{rcl}
    \mathit{enum}(\onet + (\onet + \onet)) &=& [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
    \mathit{enum}((\onet + \onet) + \onet) &=& [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
\end{array}\]
Thus, as shown in the diagrams below, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of
$\inlv{\Acon{tt}}$ which is 0 and produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value
$\inlv{(\inlv{\Acon{tt}})}$. Similarly, $\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of
$\inlv{\Acon{tt}}$ which is 0 and produces index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value
$\inrv{\Acon{tt}}$.

\begin{center}
  \input{assoc-perm.tikz}
  \qquad
  \input{swap-perm.tikz}
\end{center}

We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
$\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:

\[\begin{array}{rcll}
\sem{A} &=& (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
\end{array}\]

\noindent The denotation $\sem{c}$ of a combinator $c : A \isot B$ is a path between $\sem{A}$ and $\sem{B}$. If the
size of $A$ is $m$ and the size of $B$ is $n$, the desired path is between $(\mathsf{Fin}~m, [ m , \mathsf{refl} ])$ and
$(\mathsf{Fin}~n, [ n , \mathsf{refl} ])$. This path is directly constructed using $\mathit{ap}$ and the fact that $m=n$
since combinators are always between types of the same size.

\noindent Finally, given two combinators $p , q : A \isot_1 B$ and a 2-combinator $\alpha : p \Iso q$, the denotation
$\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.

\note{We use the rig structure of $\UFin$ in~\cref{subsec:rig} to interpret $\PiLang$.}

We need a formal definition of normal form (canonical form)

Recalling that the $\lambda$-calculus arises as the internal language of Cartesian Closed Categories
(Elliott~\cite{Elliott-2017} gives a particularly readable account of this), we can think of $\Pi$ in similar terms, but
for symmetric Rig Groupoids instead. For example, we can ask what does the equivalence above represent? It is actually a
``linear'' representation of a 2-categorial commutative diagram! In fact, it is a painfully verbose version thereof, as
it includes many \emph{refocusing} steps because our language does not build associativity into its syntax. Categorical
diagrams usually do.  Thus if we rewrite the example in diagrammatic form, eliding all uses of associativity, but
keeping explicit uses of identity transformations, we get that \AgdaFunction{swap{-}fl2⇔swap{-}fl1} represents

\vspace*{3mm}
\begin{tikzcd}[column sep=normal, row sep=normal]
 && (a+c)+b \arrow [r, "\swapp \oplus\idd", ""{name=U, below}] & (c+a)+b \arrow [dr, "\assocrp"] && \\
 & a+(c+b) \arrow [ur, "\assoclp"] & & & c+(a+b) \arrow [dr, "\idd\oplus\swapp"] &  \\
a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [r, "\assoclp"]
  \arrow [dr, "\assoclp"]
  \arrow [ddr, swap, "\assoclp"]
    & (a+b)+c \arrow [r, "\swapp"] &
    c+(a+b) \arrow [r, swap, "\assoclp", ""{name=D, above}]
    & |[alias=Z]| (c+a)+b \arrow [r, "\assocrp"] &c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
 & (a+b)+c \arrow [dr, "\swapp"] &&&& \\
 & (a+b)+c \arrow [dr, swap, "\swapp"] & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
             \arrow [d, Rightarrow, "\idf\, \mathit{idl}\odot{l}"] &&
    c+(a+b) \arrow [ruu, "\idd\oplus\swapp"] & \\
 && c+(a+b) \arrow [rrruuu, bend right = 40, swap, "\idd\oplus\swapp"] && \\
 \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
 \arrow[Rightarrow, from=Z, to=DD, swap, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
\end{tikzcd}
